Correctness of Algebraic File Synchronization

### 2015-11-25 2015/paper.tex
Finalizing text and references

\newtheorem{myax}{Axiom}
\crefname{myax}{axiom}{axioms}
% \creflabelformat{myax}{A#2#1#3}

\newtheorem{mylem}{Lemma}
\crefname{mylem}{lemma}{lemmas}
% \creflabelformat{mylem}{L#2#1#3}

@@

\newtheorem{myrul}{Rule}
\crefname{myrul}{rule}{rules}
% \creflabelformat{myrul}{R#2#1#3}

@@

We model a filesystem using a function $\FS$ with a set of nodes (filesystem paths) $\setn$ as its domain,
and a set of possible contents $\setv$ as its codomain:
\[ \FS : \setp \rightarrow \setv \] 
In our model, $\setn$ contains all possible nodes, including the ones where there is no file or directory,
and $\setv$ contains a special element, $\empt$, which is the value of $\FS$ at an empty node.
We consider the contents of files, as well as any meta-information of files
and directories (e.g. permissions or flags) part of the values in $\setv$.

@@

\begin{mydef}{$n\unrel m$.}
We write $n\unrel m$, or ``$n$ and $m$ are unrelated'' if $n\not\descendant m$ and $n\not\ancestor m$
and $n\new m$.
\end{mydef}

@@

\begin{mydef}{Minimal set or sequence.}
% TODO Don't introduce?
A sequence or set of commands is minimal if it contains at most one command on every node.
\end{mydef}

@@

From \cref{ax_separate_commute} we know that commands from different 
components can be applied in any order.
Therefore once we have ordered the commands inside a component, we generate
all permutations of the components themselves, including the ones
in which commands from separate components mix and overlap
(but where the possible orders determined for the components are respected).

@@

\begin{mylem}\label{simple_is_max}
For any simple sequence $S\neq\cbrk$, $\forall T: S\eqext T \Rightarrow S\equiv T$;
that is, the domains of simple sequences cannot be extended further.
This is equivalent to $\extset{S} = \{S'|S\equiv S'\}$.
\end{mylem}
\begin{proof}
TODO
% TODO
\end{proof}

@@

\begin{mydef}{Independent commands, sequences and sets.}
For two commands $\cxy$ and $\czw$ we write $\cxy\indep\czw$ if $\cxy\cc \czw\equiv\czw\cc \cxy\nequiv\cbrk$.
For two sequences or sets of commands $A$ and $B$ we write $A\indep B$ if
$\forall a\in A \mbox{and~} \forall b\in B: a\indep b$.
\end{mydef}

@@

In other words, the sequence $\reca$ we apply to $\FS_B$ consists of all commands from $A\setminus B$
that are independent of every command in $B\setminus A$.
The order of commands in the sequence can be determined using the algorithm described in
\cref{ordering} as any subset of $A$ is also simple.
\end{mydef}

Now we prove that $\reca\FS_B\neq\fsbroken$, and, in general:

% 'Works'
% -------

\begin{mydef}{Inverse commands and sequences.}
The inverse of command $\cxy(p,V_Y)$ is $\cxy(p,V_Y)^{-1} = \cyx(p,V_X)$
where $V_X$ is an arbitrary value from $\setvx{X}$.
The inverse of sequence $S$, $S^{-1}$, consists of the inverses of the commands in $S$
in reverse order.
\end{mydef}

\begin{mylem}\label{r_invmove}
\[ \infer{\worksc{S\cc A}{B}}{\worksc{A}{S^{-1}\cc B}} \]
To see the soundness of this rule...
Sequences are bijections on filesystems over type-equality.
TODO: $A,S,B$ need to be simple?
\end{mylem}

\begin{myrul}\label{r_indep_combine}
\[ \inferrr{\worksc{T}{S}}{\worksc{U}{S}}{T\indep U}{\worksc{T\cc U}{S}} \]
To see the soundness of this rule...
\end{myrul}

\begin{mylem}\label{indep_prefix_combine}
\[ \inferrr{\worksc{A\cc B}{S}}{\worksc{A\cc C}{S}}{B\indep C}{\worksc{A\cc B\cc C}{S}} \]
\end{mylem}
\begin{proof}
From \cref{r_invmove,r_indep_combine}
TODO reverse of \cref{r_invmove} needed
\end{proof}

\begin{myax}\label{worksextpostfix}
$\worksc{A}{A\cc S}$ where $A$ and $S$ are arbitrary sequences of commands
\end{myax}

% Introducing 'works'
% -------------------

So that we can formalise statements (TODO) needed to prove this,
let us introduce two relations that describe under what conditions
sequences of commands work, that is, do not break a filesystem.

\begin{mydef}
% Unsure if this relation should be in the algebra.
For a set of $k$ sequences
$\works{A_1,A_2,\ldots,A_k}$ means that 
$A_1,A_2,\ldots,A_k$ work at the same time, that is,
% intended interpretation:
\[\exists \FS: A_1\FS\neq\fsbroken \wedge \cdots \wedge A_k\FS\neq\fsbroken\]
As the sequences form a set, their order is irrelevant.
% inference rules:
We also see that $\works{A} \Leftrightarrow A\nequiv \cbrk$. 
\end{mydef}

\begin{mydef}
For two sets of sequences, $\worksc{A_1,A_2,\ldots,A_k}{B_1,B_2,\ldots,B_l}$ means that 
all of $A_1,A_2,\ldots,A_k$ work where all of $B_1,B_2,\ldots,B_l$ work,
that is,
% intended interpretation:
\begin{align*}
\forall \FS:&\\ 
&B_1\FS\neq\fsbroken \wedge \cdots \wedge B_k\FS\neq\fsbroken\\
&\Rightarrow\\
&A_1\FS\neq\fsbroken \wedge \cdots \wedge A_l\FS\neq\fsbroken
\end{align*}
As above, the order of the sequences in the sets is not relevant.
% inference rules:
It is easy to see that 
% $\worksc{X}{X}$ is always true for any set of sequences $X$,
$B\eqext A \Rightarrow \worksc{A}{B}$
and that this relation can be chained:
$ \worksc{A_1,\ldots,A_k}{B_1,\ldots,B_l} \wedge \worksc{C_1,\ldots,C_j}{A_1} \Rightarrow \worksc{A_1,\ldots,A_k,C_1,\ldots,C_j}{B_1,\ldots,B_l} $
\end{mydef}

@@

\begin{myth}
If $A, B$ are simple, then $\worksc{B\cc \reca}{A,B}$
\end{myth}

This is trivial unless $\works{A,B}$.

\begin{mylem}
Let $A$ and $B$ be two simple sequences so that $\works{A, B}$.
Then their intersection works on any filesystem
either of them works on:
$\worksc{\ordered{A\cap B}}{A}$, and, by symmetry, for $B$.
\end{mylem}

\begin{proof}
First we prove the following.
Mark all commands in $A$ that also appear in $B$.
Then there is an $A'$ in $\orderset{A}$ in which the marked commands are at the beginning.

@@

TODO: unmarked-marked can be reversed -- but induction is wrong

We show $A$ can be transformed via equivalences so that this would be true,
by induction on the number of non-marked commands preceding the last marked command.
If it is 0, no transformation is needed.
For the induction step, consider the last marked command preceded by a non-marked command,
and the non-marked command itself.
As $A$ is a simple command and $\works{A}$, from 
\cref{ax_distantchild_breaks,ax_distantparent_breaks,ax_directchild_breaks,ax_directparent_breaks}
we know that these commands can only be on unrelated paths or form a construction or destruction pair.
In the first case, swapping the commands results in a sequence equivalent to $A$,
with the number on relevant non-marked commands reduced by one.
In the last two cases,
either $q=\parent(p)$ or $p=\parent(q)$.

Let the marked command be $\cxy(p, V_Y)$ and the preceding unmarked command in $A$ be $\czw(q, V_W)$.

If $B$ has a command on $q$, then
from \cref{lemma:neighbor}
we know that there is a $B'$ in $\orderset{B}$ where it is next to $\cxy$,
and from \cref{lemma:minextset,lemma:minextorder} we know that $B'\equiv B$.
% TODO Use when describing ordering:
Considering the eight construction and destruction pairs we see that the command
on the parent path determines which group the pair belongs to.
This in turn determines whether the command on the child path must
precede or follow the other command as otherwise the sequence would break all filesystems.
This argument holds for both $A$ and $B$, and so the command on $q$ must be on
the same side of the command on $p$ in both sequences.

If $B$ has no command on $q$, then let $B'$ be $B$, but with an extra command added to it just before $\cxy$
according to \cref{ax_child_assert,ax_parent_assert}, from which we know
that $B'\equiv B$.
We also know that $B'$ is still minimal (but no longer simple).

We therefore have a minimal $B'$ which has a command on $q$ just before $\cxy(p, V_Y)$ and $B'\equiv B$.
Let this command on $q$ be $\cqr(q, V_R)$.
\begin{gather*}
A = \cdots\cc  \czw(q, V_W)\cc  \cxy(p, V_Y)\cc  \cdots \\
B' = \cdots\cc  \cqr(q, V_R)\cc  \cxy(p, V_Y)\cc  \cdots
\end{gather*}

As $\works{A,B}$ and so $\works{A,B'}$ (via \cref{r_eq_ex,r_ex_w,r_w_w}), 
from \cref{worksinputmatch}
we know that $Q=Z$.

Going back to the construction and destruction pairs we see that the output type of the first command
is always the same, and it is either $D$ or $\empt$ depending on the relationship between $p$ and $q$.
Therefore $R=W$ and $\cqr(q, V_R)=\czw(q, V_W)$
as $|\setd|=|\setb|=1$. 
If $B$ originally had a command on $q$,
this is a contradiction as $\czw$ was not marked, but we see it must also be in $B'$ and therefore in $B$.
If $B$ had no command on $q$,
this is a contradiction because the command on $q$ in $B'$ is an assertion command, so $Z=Q=R=W$, 
but $A$ contained no assertion commands.

@@

We now know that there is an $A'\equiv A$ in which commands in $A\cap B$
are at the beginning, and therefore 
from \cref{worksextpostfix}
$\worksc{\ordered{A\cap B}}{A}$ and by symmetry $\worksc{\ordered{A\cap B}}{B}$.
\end{proof}

@@

We would like to prove that $\worksc{B\cc \reca}{A,B}$.
From above we know that this is equivalent to
$\worksc{\ordered{A\cap B}\cc \ordered{B\setminus A}\cc \reca}{A,B}$.
As we already know that $\reca\indep B\setminus A$
and that $\worksc{\ordered{A\cap B}\cc \ordered{B\setminus A}}{A,B}$
(as $\worksc{B}{A,B}$),
based on \cref{indep_prefix_combine}
we only need to prove that
$\worksc{\ordered{A\cap B}\cc \reca}{A,B}$.

\newcommand{\acbi}{\ordered{A\cap B}^{-1}}
\newcommand{\amb}{A\setminus B}
\newcommand{\bma}{B\setminus A}
Using \cref{r_invmove} we can rephrase this as
% TODO axiom doesn't specify multiple conditions
$\worksc{\reca}{\acbi\cc A,\acbi\cc B}$.

@@

Similarly to above, we prove that $S$ can be transformed into $S'$ by equivalences
so that all commands in $S^*$ would be at its beginning.
If so then based on \cref{worksextpostfix}
$\worksc{S^*}{C_1,\ldots,C_n}$ as $\worksc{S'}{C_1,\ldots,C_n}$.

>>>>

\newtheorem{myax}{Axiom}
\crefname{myax}{axiom}{axioms}

\newtheorem{mylem}{Lemma}
\crefname{mylem}{lemma}{lemmas}

@@

\newtheorem{mycor}{Corollary}
\crefname{mycor}{corollary}{corollaries}

% TODO delete:
\newtheorem{myrul}{Rule}
\crefname{myrul}{rule}{rules}

@@

We model a filesystem using a function $\FS$ with a set of nodes (filesystem paths) $\setn$ as its domain,
and a set of possible contents $\setv$ as its codomain:
\[ \FS : \setn \rightarrow \setv \] 
In our model, $\setn$ contains all possible nodes, including the ones where there is no file or directory,
and $\setv$ contains a special element, $\empt$, which is the value of $\FS$ at an empty node.
We consider the contents of files, as well as any meta-information of files
and directories (e.g. permissions or flags) part of the values in $\setv$.

@@

\begin{mydef}{$n\unrel m$.}
We write $n\unrel m$, or ``$n$ and $m$ are unrelated'' if $n\not\descendant m$ and $n\not\ancestor m$
and $n\neq m$.
\end{mydef}

@@

\begin{mydef}{Minimal set or sequence.}
A sequence or set of commands is minimal if it contains at most one command on every node.
\end{mydef}

@@

From \cref{ax_separate_commute} we know that commands from different 
components can be applied in any order.
Therefore once we have ordered the commands inside a component, we generate
all permutations of the components themselves, including the ones
in which commands from separate components mix and overlap
(but where the possible orders determined for commands inside the components are respected).

@@

\begin{mylem}\label{simple_is_max}
For any simple sequence $S\neq\cbrk$, $\forall T: S\eqext T \Rightarrow S\equiv T$;
that is, the domains of simple sequences cannot be extended further.
This is equivalent to the proposition that $\extset{S} = \{S'|S\equiv S'\}$.
\end{mylem}
\begin{proof}
TODO
% TODO
\end{proof}

@@

\begin{mydef}{Independent commands, sequences and sets.}
For two commands $\cxy$ and $\czw$ we write $\cxy\indep\czw$ if $\cxy\cc \czw\equiv\czw\cc \cxy\nequiv\cbrk$.
For two sequences or sets of commands $A$ and $B$ we write $A\indep B$ if
$\forall a\in A \mbox{~and~} \forall b\in B: a\indep b$.
\end{mydef}

@@

In other words, the sequence $\reca$ we apply to $\FS_B$ consists of all commands from $A\setminus B$
that are independent of every command in $B\setminus A$.
The order of commands in the sequence can be determined using the algorithm described in
\cref{ordering} as any subset of $A$ is also simple,
and $\recb$ can be obtained in a similar way by reversing $A$ and $B$.
\end{mydef}

We would now like to prove that $\reca$ and $\recb$ can be applied to the replicas,
that is, without loss of generality,
\[ \reca\FS_B\neq\fsbroken. \]
So that we can formalise statements needed to prove this,
let us introduce two relations that describe under what conditions
sequences of commands work, that is, do not break a filesystem.

% Works
% -----

\begin{mydef}
% Unsure if this relation should be in the algebra.
For a set of $k$ sequences
$\works{A_1,A_2,\ldots,A_k}$ means that 
$A_1,A_2,\ldots,A_k$ work at the same time, that is,
% intended interpretation:
\[\exists \FS: A_1\FS\neq\fsbroken \wedge \cdots \wedge A_k\FS\neq\fsbroken\]
As the sequences form a set, their order is irrelevant.
% inference rules:
We also know that $\works{A} \Leftrightarrow A\nequiv \cbrk$. 
\end{mydef}

\begin{mydef}
For two sets of sequences, $\worksc{A_1,A_2,\ldots,A_k}{B_1,B_2,\ldots,B_l}$ means that 
all of $A_1,A_2,\ldots,A_k$ work where all of $B_1,B_2,\ldots,B_l$ work,
that is,
% intended interpretation:
\begin{align*}
\forall \FS:&\\ 
&B_1\FS\neq\fsbroken \wedge \cdots \wedge B_k\FS\neq\fsbroken\\
&\Rightarrow\\
&A_1\FS\neq\fsbroken \wedge \cdots \wedge A_l\FS\neq\fsbroken
\end{align*}
As above, the order of the sequences in the sets is not relevant.
\end{mydef}

% An inference rule in the algebra
% $\worksc{X}{X}$ is always true for any set of sequences $X$,

It is easy to see that the following corollaries are true:

\begin{mycor}
% An inference rule in the algebra
$B\eqext A \Rightarrow \worksc{A}{B}$
\end{mycor}

\begin{mycor}\label{worksextpostfix}
% An inference rule in the algebra
$\forall A,S: \worksc{A}{A\cc S}$
\end{mycor}

\begin{mycor}
% An inference rule in the algebra
$\worksc{S}{T}$ can be chained:
$ \worksc{A_1,\ldots,A_k}{B_1,\ldots,B_l} \wedge \worksc{C_1,\ldots,C_j}{A_1} \Rightarrow \worksc{A_1,\ldots,A_k,C_1,\ldots,C_j}{B_1,\ldots,B_l} $
\end{mycor}

@@

\begin{mylem}\label{r_indep_combine}
% An inference rule in the algebra
\[ \worksc{T}{S} \wedge \worksc{U}{S} \wedge T\indep U \Rightarrow \worksc{T\cc U}{S} \]
\end{mylem}
\begin{proof}
% TODO Proof
TODO
\end{proof}

@@

We continue by defining inverse commands and sequences
which allow us to move parts of sequences between the
two sets of $\worksc{S}{T}$.

\begin{mydef}{Inverse commands and sequences.}
The inverse of command $\cxy(n,V_Y)$ is $\cxy(n,V_Y)^{-1} = \cyx(n,V_X)$
where $V_X$ is an arbitrary value from $\setvx{X}$.
We write $S^{-1}$ for the inverse of sequence $S$, which consists of the inverses of the commands in $S$
in reverse order.
%
Clearly, $(S^{-1})^{-1}\equiv S$ for any sequence.
\end{mydef}

\begin{mylem}\label{r_invmove}
\begin{gather*}
\worksc{B\cc A}{S} \Rightarrow \worksc{A}{B^{-1}\cc S} \\
\worksc{A}{B\cc S} \Rightarrow \worksc{B^{-1}\cc A}{S}
\end{gather*}
% TODO Proof
TODO.
Sequences are bijections on filesystems over type-equality.
TODO: $A,S,B$ need to be simple?
\end{mylem}

\begin{mylem}\label{indep_prefix_combine}
\[ \worksc{A\cc B}{S} \wedge \worksc{A\cc C}{S} \wedge B\indep C \Rightarrow \worksc{A\cc B\cc C}{S} \]
\end{mylem}
\begin{proof}
Based on \cref{r_invmove} we know
$\worksc{B}{A^{-1}\cc S}$ and $\worksc{C}{A^{-1}\cc S}$,
and from \cref{r_indep_combine} we know
$\worksc{B\cc C}{A^{-1}\cc S}$. Applying the second line of \cref{r_invmove} we get
$\worksc{(A^{-1})^{-1}\cc B\cc C}{S}$, which proves our lemma as $(A^{-1})^{-1}\equiv A$.
\end{proof}

@@

% Correctness of reconciliation
% -----------------------------

We are now ready to prove that the proposed algorithm for reconciliation is correct,
that is, applying its result is not going to break the replicas.
We reformulate the original proposition, $\reca\FS_B\neq\fsbroken$,
based on $\FS_B=B\FS$ and so we aim to prove that
$B\cc\reca$ works wherever $A$ and $B$ work:

\begin{myth}
If $A, B$ are simple, then $\worksc{B\cc \reca}{A,B}$
\end{myth}
where, to restate the definition,
{ % scope
\renewcommand{\themydef}{\ref{def:reconciliation}}
\begin{mydef}
\begin{align*}
\reca = \{c| & \\
& c\in A\setminus B \\
& \wedge \\
& \forall c'\in B\setminus A: c\indep c'\\
\}. &
\end{align*}
\end{mydef}
} % scope

This is trivial unless $\works{A,B}$, so we assume that it is true.
Let us first investigate the part of $A$ and $B$ that is excluded from
$\reca$: their intersection.

\begin{mylem}
Let $A$ and $B$ be two simple sequences so that $\works{A, B}$.
Then their intersection works on any filesystem
either of them works on:
$\worksc{A\cap B}{A}$, and, by symmetry,
$\worksc{A\cap B}{B}$.
\end{mylem}

\begin{proof}
First we prove the following.
Mark all commands in $A$ that also appear in $B$.
Then there is an $A'$ in $\orderset{A}$ in which all marked commands are at the beginning.
We show $A$ can be transformed via equivalences so that this would be true.
Our proposition is that if the marked commands are not at the beginning, then
the sequence contains an unmarked command followed by a marked command.
We show that these can be reversed resulting in an equivalent sequence.
Then, by repeating this process similarly to bubble sorting, we can generate 
a suitable $A'$.

@@

Let us consider therefore the marked command preceded by an unmarked command in $A$.
and let the marked command be $\cxy(n, V_Y)$, 
and the preceding unmarked command be $\czw(m, V_W)$.
As $A$ is a simple command and $\works{A}$, from 
\cref{ax_distantrel_breaks,ax_directchild_breaks,ax_directparent_breaks}
we know that these commands can only be on unrelated paths or form a construction or destruction pair.
In the first case, swapping the commands results in a sequence equivalent to $A$,
and we show that the last two cases are impossible as they would lead to contradiction.

In these cases, either $m=\parent(n)$ or $p=\parent(m)$.
If $B$ has a command on $m$, then
from \cref{lemma:neighbor}
we know that there is a $B'$ in $\orderset{B}$ where it is next to $\cxy$ (also in $B$),
and from \cref{simple_is_max,lemma:minextorder} we know that $B'\equiv B$.
When describing our ordering algorithm in \cref{ordering} we saw
that the command on the parent path deremines whether a pair of commands
is a construction or destruction pair,
and that this, in turn, determines whether the command on the child path must
precede or follow the other command as otherwise the sequence would break all filesystems.
This argument holds for both $A$ and $B$, and so the command on $m$ must be on
the same side of the command on $n$ in both sequences.

If $B$ has no command on $m$, then let $B'$ be $B$, 
but with an extra command added to it just before $\cxy$
according to \cref{ax_child_assert,ax_parent_assert}, 
from which we know that $B'\equiv B$.
We also know that $B'$ is still minimal (but no longer simple).

In either case, therefore,
we have a minimal $B'$ which has a command on $m$ just before $\cxy(n, V_Y)$ and $B'\equiv B$.
Let this command on $m$ be $\cqr(m, V_R)$.
\begin{gather*}
A = \cdots\cc  \czw(q, V_W)\cc  \cxy(p, V_Y)\cc  \cdots \\
B' = \cdots\cc  \cqr(q, V_R)\cc  \cxy(p, V_Y)\cc  \cdots
\end{gather*}

As $\works{A,B}$ and so $\works{A,B'}$, 
from \cref{worksinputmatch}
we know that $Q=Z$.

Going back to the construction and destruction pairs we see that the output type of the first command
is always the same, and it is either $D$ or $\empt$ depending on the relationship between $n$ and $m$.
Therefore $R=W$ and $\cqr(m, V_R)=\czw(m, V_W)$
as $|\setd|=|\setb|=1$. 
If $B$ originally had a command on $m$,
this is a contradiction as $\czw$ was not marked, but we see it must also be in $B'$ and therefore in $B$.
If $B$ had no command on $m$,
this is a contradiction because the command on $m$ in $B'$ is an assertion command, so $Z=Q=R=W$, 
but $A$ contained no assertion commands.

We now know that there is an $A'\equiv A$ in which commands in $A\cap B$
are at the beginning, and therefore 
from \cref{worksextpostfix}
$\worksc{A\cap B}{A}$ and by symmetry $\worksc{A\cap B}{B}$.
\end{proof}

We would like to prove that $\worksc{B\cc \reca}{A,B}$.
From the above we know that this is equivalent to
$\worksc{(A\cap B)\cc (B\setminus A)\cc \reca}{A,B}$.
As we already know that $\reca\indep (B\setminus A)$
and that $\worksc{(A\cap B)\cc(B\setminus A)}{A,B}$
(as $\worksc{B}{A,B}$),
based on \cref{indep_prefix_combine}
we only need to prove that
$\worksc{(A\cap B)\cc \reca}{A,B}$.

% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\newcommand{\acbi}{(A\cap B)^{-1}}
\newcommand{\amb}{(A\setminus B)}
\newcommand{\bma}{(B\setminus A)}
Using \cref{r_invmove} we can rephrase this as
% TODO axiom doesn't specify multiple conditions
$\worksc{\reca}{\acbi\cc A,\acbi\cc B}$.

@@

Similarly to above, we prove that $S$ can be transformed into $S'$ by equivalences
so that all commands in $S^*$ would be at its beginning.
If so, then based on \cref{worksextpostfix}
$\worksc{S^*}{C_1,\ldots,C_n}$ as $\worksc{S'}{C_1,\ldots,C_n}$ and .
